---
title: "Wearable Device Comparative Analysis"
subtitle: "Whoop vs Oura Ring 4 vs Withings ScanWatch"
author: "Wearable Comparative Analysis"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: false
    theme: cosmo
execute:
  echo: true
  message: false
  warning: false
---

```{r setup}
# Set working directory to project root so sourced scripts find data files.
knitr::opts_chunk$set(fig.width = 9, fig.height = 5, dpi = 120)

if (!file.exists("physiological_cycles.csv") && file.exists("../physiological_cycles.csv")) {
  setwd("..")
}

source("R/00_packages.R")
source("R/source_project.R")
source("R/06_plots.R")

# Run the full pipeline: load, clean, harmonize, save processed files.
harmonized <- save_harmonized()
long_df  <- harmonized$long
wide_df  <- harmonized$wide
overlap  <- harmonized$overlap
qc_log   <- harmonized$qc

desc_stats    <- descriptive_stats(long_df)
agreement     <- run_all_agreement(wide_df)
overlap_tbl   <- overlap_counts(overlap)
ww_stats      <- weekday_weekend_stats(long_df)
```

## Introduction

This report compares daily health metrics from three wearable devices worn by the same person over approximately 33 days (June 20 – July 22, 2026, US Pacific time):

- **Whoop** bracelet
- **Oura Ring 4**
- **Withings ScanWatch**

We examine four metrics: **resting heart rate (RHR)**, **heart rate variability (HRV)**, **sleep duration**, and **daily steps**. The analysis window is anchored to Whoop coverage. All code is shown inline so the workflow is fully transparent.

## Data sources and analysis window

| Device | Primary files | Metrics extracted |
|--------|---------------|-------------------|
| Whoop | `physiological_cycles.csv`, `whoop.rtf` | RHR, HRV, sleep; steps manually logged |
| Oura Ring 4 | `oura_2026-05-22_2026-07-23_trends.csv` | RHR, HRV, sleep, steps |
| Withings ScanWatch | `sleep.csv`, `aggregates_steps.csv` | RHR (nightly avg), sleep, steps; **HRV not available in 2026 export** |

```{r window-info}
window_info <- tibble::tibble(
  Setting = c("Timezone", "Window start", "Window end", "Calendar days"),
  Value = c(TZ, as.character(WINDOW_START), as.character(WINDOW_END),
            as.character(as.integer(WINDOW_END - WINDOW_START) + 1))
)
knitr::kable(window_info, caption = "Analysis configuration")
```

### Sleep date alignment

All three devices naturally align to the **wake date in Pacific time**:

| Device | Rule |
|--------|------|
| Whoop | Date of `Wake onset` |
| Oura | `date` column (summary / wake calendar day) |
| Withings | Date of sleep end timestamp (`to`) |

## Data quality and exclusions

Known issues handled in the pipeline:

- Whoop steps are **manually transcribed** from the app (not exported)
- Withings sleep on 2026-06-20 is corrupt (zero stages) — excluded
- Withings duplicate sleep on 2026-06-26 — kept longest session
- Withings HRV absent in 2026 — HRV comparisons are Whoop vs Oura only
- Withings sleep data ends 2026-07-18 (no Jul 19–22 nights)
- Low step counts flagged (not dropped) for sensitivity analysis

```{r qc-table}
if (nrow(qc_log) > 0) {
  knitr::kable(head(qc_log, 20), caption = "QC exclusions and flags (first 20 rows)")
} else {
  cat("No hard exclusions beyond missing values.")
}
```

## Overlap and coverage

How many days does each device contribute per metric?

```{r overlap-table}
knitr::kable(overlap_tbl, caption = "Days by device-combination category")
```

```{r overlap-heatmaps, fig.height=8}
#| fig-cap: "Calendar heatmaps showing which devices recorded each metric on each day."
plots <- lapply(METRICS, function(m) plot_overlap_heatmap(overlap, m))
patchwork::wrap_plots(plots, ncol = 1)
```

## Descriptive statistics

```{r desc-table}
desc_display <- desc_stats |>
  dplyr::mutate(
    metric = METRIC_LABELS[as.character(metric)],
    dplyr::across(where(is.numeric), ~ round(.x, 2))
  ) |>
  dplyr::select(device, metric, n, mean, sd, median, q25, q75, cv_pct)

knitr::kable(desc_display, caption = "Summary statistics by device and metric")
```

```{r timeseries, fig.height=10}
#| fig-cap: "Daily time series for each metric, colored by device."
ts_plots <- lapply(METRICS, function(m) plot_timeseries(long_df, m))
patchwork::wrap_plots(ts_plots, ncol = 2)
```

```{r density, fig.height=10}
#| fig-cap: "Distribution of daily values per device."
dens_plots <- lapply(METRICS, function(m) plot_density(long_df, m))
patchwork::wrap_plots(dens_plots, ncol = 2)
```

## Correlation and agreement

Pairwise comparisons on days where both devices recorded the metric. HRV includes only Whoop and Oura (Withings has no HRV in this export).

```{r agreement-table}
if (!is.null(agreement)) {
  agree_display <- agreement |>
    dplyr::mutate(
      metric = METRIC_LABELS[metric],
      dplyr::across(where(is.numeric), ~ round(.x, 3))
    )
  knitr::kable(agree_display, caption = "Pairwise agreement statistics")
}
```

```{r corr-heatmap, fig.height=5}
#| fig-cap: "Pearson correlation heatmap across device pairs."
if (!is.null(agreement)) plot_correlation_heatmap(agreement)
```

### Scatter plots and Bland–Altman diagrams

```{r scatter-ba-rhr, fig.height=10}
#| fig-cap: "RHR: scatter and Bland–Altman plots for each device pair."
p1 <- plot_scatter_pair(wide_df, "rhr", "whoop", "oura", "Whoop", "Oura Ring 4")
p2 <- plot_bland_altman(wide_df, "rhr", "whoop", "oura", "Whoop", "Oura Ring 4")
p3 <- plot_scatter_pair(wide_df, "rhr", "whoop", "withings", "Whoop", "Withings ScanWatch")
p4 <- plot_bland_altman(wide_df, "rhr", "whoop", "withings", "Whoop", "Withings ScanWatch")
patchwork::wrap_plots(list(p1, p2, p3, p4), ncol = 2)
```

```{r scatter-ba-sleep, fig.height=10}
#| fig-cap: "Sleep duration: scatter and Bland–Altman plots."
s1 <- plot_scatter_pair(wide_df, "sleep_min", "whoop", "oura", "Whoop", "Oura Ring 4")
s2 <- plot_bland_altman(wide_df, "sleep_min", "whoop", "oura", "Whoop", "Oura Ring 4")
s3 <- plot_scatter_pair(wide_df, "sleep_min", "whoop", "withings", "Whoop", "Withings ScanWatch")
s4 <- plot_bland_altman(wide_df, "sleep_min", "whoop", "withings", "Whoop", "Withings ScanWatch")
patchwork::wrap_plots(list(s1, s2, s3, s4), ncol = 2)
```

```{r scatter-ba-steps, fig.height=10}
#| fig-cap: "Steps: scatter and Bland–Altman plots (Whoop steps are manually logged)."
t1 <- plot_scatter_pair(wide_df, "steps", "whoop", "oura", "Whoop", "Oura Ring 4")
t2 <- plot_bland_altman(wide_df, "steps", "whoop", "oura", "Whoop", "Oura Ring 4")
t3 <- plot_scatter_pair(wide_df, "steps", "oura", "withings", "Oura Ring 4", "Withings ScanWatch")
t4 <- plot_bland_altman(wide_df, "steps", "oura", "withings", "Oura Ring 4", "Withings ScanWatch")
patchwork::wrap_plots(list(t1, t2, t3, t4), ncol = 2)
```

```{r scatter-ba-hrv, fig.height=5}
#| fig-cap: "HRV: Whoop vs Oura only (Withings HRV unavailable)."
h1 <- plot_scatter_pair(wide_df, "hrv", "whoop", "oura", "Whoop", "Oura Ring 4")
h2 <- plot_bland_altman(wide_df, "hrv", "whoop", "oura", "Whoop", "Oura Ring 4")
patchwork::wrap_plots(list(h1, h2), ncol = 2)
```

## Extended analyses

### Weekday vs weekend

```{r weekday-weekend}
knitr::kable(
  ww_stats |> dplyr::mutate(metric = METRIC_LABELS[as.character(metric)],
                            mean = round(mean, 2)),
  caption = "Mean values on weekdays vs weekends"
)
```

### Rank consistency

On days where all three devices recorded a metric, do they rank days similarly (Spearman correlation of ranks)?

```{r rank-consistency}
rank_rhr   <- rank_consistency(wide_df, "rhr", c("whoop", "oura", "withings"))
rank_sleep <- rank_consistency(wide_df, "sleep_min", c("whoop", "oura", "withings"))
rank_steps <- rank_consistency(wide_df, "steps", c("whoop", "oura", "withings"))
rank_all <- dplyr::bind_rows(rank_rhr, rank_sleep, rank_steps)

if (!is.null(rank_all) && nrow(rank_all) > 0) {
  knitr::kable(
    rank_all |> dplyr::mutate(metric = METRIC_LABELS[metric], spearman = round(spearman, 3)),
    caption = "Rank correlation between devices (Spearman)"
  )
}
```

## Limitations

- **Single participant**, ~33-day window — correlations are exploratory, not population-level inference.
- **Whoop steps** were manually typed into an RTF note; gaps and low values reduce reliability.
- **Withings HRV** is not present in the 2026 export; only Whoop–Oura HRV comparison is possible.
- **Withings sleep** stops after July 18, shortening three-way sleep overlap.
- Devices use **proprietary algorithms**; disagreement does not imply clinical inaccuracy without a reference standard (e.g., polysomnography, ECG).
- All timestamps interpreted in **`r TZ`**.

## Appendix

```{r session-info}
sessionInfo()
```
