---
title: "Wearable Device Comparative Analysis"
subtitle: "Whoop vs Oura Ring 4 vs Withings ScanWatch"
author: "Wearable Comparative Analysis"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: true
    theme: cosmo
    embed-resources: true
execute:
  echo: true
  message: false
  warning: false
---

```{r setup}
#| code-fold: true
knitr::opts_chunk$set(fig.width = 9, fig.height = 5, dpi = 120)

if (!file.exists("physiological_cycles.csv") && file.exists("../physiological_cycles.csv")) {
  setwd("..")
}

source("R/00_packages.R")
source("R/source_project.R")
source("R/06_plots.R")

harmonized <- save_harmonized()
long_df    <- harmonized$long
wide_df    <- harmonized$wide
overlap    <- harmonized$overlap
qc_log     <- harmonized$qc

# Core-metric agreement (headline four).
agreement      <- run_all_agreement(wide_df, metrics = METRICS)
# Extended metrics (stages, timing, rhr_rest, etc.).
agreement_ext  <- run_all_agreement(wide_df, metrics = setdiff(ALL_METRICS, METRICS))

desc_stats     <- descriptive_stats(long_df)
overlap_tbl    <- overlap_counts(overlap)
ww_stats       <- weekday_weekend_stats(long_df)
exec_summary   <- executive_summary(agreement)
compliance     <- compliance_summary(long_df)
kendall_tbl    <- kendall_w_table(wide_df)
arch_df        <- sleep_architecture_wide(long_df)
workouts       <- parse_whoop_workouts()
intraday_hr    <- load_withings_intraday_hr()

# Sensitivity: re-run without QC-flagged days.
sensitivity    <- sensitivity_agreement(long_df, wide_df, metrics = METRICS)
```

## Executive summary

This report compares **Whoop**, **Oura Ring 4**, and **Withings ScanWatch** over **33 days** (June 20 – July 22, 2026, US Pacific). Four headline metrics — resting heart rate, HRV, sleep duration, and steps — are analyzed with correlation, Bland–Altman agreement, and extended exploratory sections.

**Color key throughout:** <span style="color:#E63946">**Red = Whoop**</span>, <span style="color:#2A9D8F">**Green = Oura**</span>, <span style="color:#457B9D">**Blue = Withings**</span>.

```{r exec-summary}
if (!is.null(exec_summary)) {
  knitr::kable(exec_summary, caption = "Headline agreement verdicts (best pair per metric)")
}
```

**Key takeaways at a glance:**

- **Whoop ↔ Oura** agree strongly on RHR, HRV, sleep, and steps (Pearson r > 0.9 for all four).
- **Withings RHR** using average sleeping HR disagrees with Whoop/Oura (r ≈ 0.05); switching to **minimum sleeping HR** (`rhr_rest`) improves agreement substantially.
- **Sleep stages** (deep / REM / light) are available from all three devices and show moderate agreement.
- **Withings HRV** is absent in the 2026 export; HRV comparisons are Whoop–Oura only.
- **Withings sleep** ends July 18, shortening three-way sleep overlap.

## Introduction

Three consumer wearables worn concurrently by the same person:

- **Whoop** bracelet — recovery-focused; exports physiological cycles
- **Oura Ring 4** — readiness / sleep ring; daily trends export
- **Withings ScanWatch** — hybrid smartwatch; Health Mate export (3-year history, filtered here to the Whoop window)

All code is shown inline (click **Code** to expand any chunk). Pacific timezone (`America/Los_Angeles`) is used throughout; sleep is keyed to **wake date**.

## Data sources and analysis window

| Device | Primary files | Metrics extracted |
|--------|---------------|-------------------|
| Whoop | `physiological_cycles.csv`, `whoop.rtf` | RHR, HRV, sleep, stages, efficiency, respiratory rate, timing; steps (manual RTF) |
| Oura Ring 4 | `oura_2026-05-22_2026-07-23_trends.csv` | RHR, lowest RHR, HRV, sleep, stages, efficiency, respiratory rate, timing, steps |
| Withings ScanWatch | `sleep.csv`, `aggregates_steps.csv` | Sleep, stages, efficiency, avg HR, min HR, timing, steps |

```{r window-info}
window_info <- tibble::tibble(
  Setting = c("Timezone", "Window start", "Window end", "Calendar days"),
  Value = c(TZ, as.character(WINDOW_START), as.character(WINDOW_END),
            as.character(as.integer(WINDOW_END - WINDOW_START) + 1))
)
knitr::kable(window_info, caption = "Analysis configuration")
```

### Sleep date alignment

| Device | Canonical `date` rule |
|--------|----------------------|
| Whoop | Date of `Wake onset` |
| Oura | `date` column (wake / summary calendar day) |
| Withings | Date of sleep end timestamp (`to`) |

## Device compliance

How many days in the 33-day window did each device record each metric?

```{r compliance}
knitr::kable(
  compliance |>
    dplyr::mutate(
      metric = METRIC_LABELS[as.character(metric)],
      compliance_pct = round(compliance_pct, 1)
    ) |>
    dplyr::select(device, metric, days_present, compliance_pct),
  caption = "Data completeness (% of 33-day window with valid data)"
)
```

Oura has the highest overall compliance; Withings drops off for sleep after July 18 and has no HRV in 2026.

## Data quality and exclusions

```{r qc-table}
if (nrow(qc_log) > 0) {
  knitr::kable(head(qc_log, 25), caption = "QC exclusions and flags (first 25 rows)")
} else {
  cat("No hard exclusions beyond missing values.")
}
```

Handled explicitly: corrupt Withings sleep (Jun 20), duplicate session (Jun 26), Whoop manual steps with gaps, low-step off-wrist days (flagged, not dropped).

## Overlap and coverage

```{r overlap-table}
knitr::kable(overlap_tbl |> dplyr::filter(metric %in% METRICS),
             caption = "Days by device-combination (headline metrics)")
```

```{r overlap-heatmaps, fig.height=8}
#| fig-cap: "Calendar heatmaps — which devices recorded each headline metric each day."
plots <- lapply(METRICS, function(m) plot_overlap_heatmap(overlap, m))
patchwork::wrap_plots(plots, ncol = 1)
```

## Descriptive statistics

```{r desc-table}
desc_display <- desc_stats |>
  dplyr::filter(metric %in% METRICS) |>
  dplyr::mutate(
    metric = METRIC_LABELS[as.character(metric)],
    dplyr::across(where(is.numeric), ~ round(.x, 2))
  ) |>
  dplyr::select(device, metric, n, mean, sd, median, cv_pct)
knitr::kable(desc_display, caption = "Summary statistics — headline metrics")
```

Whoop and Oura track closely on RHR (~69–70 bpm) and HRV (~28–33 ms). Withings average sleeping HR (~70 bpm) is similar in mean but poorly correlated day-to-day (see RHR section below).

```{r ridgelines, fig.height=10}
#| fig-cap: "Ridgeline density plots — clearer separation than overlapping curves."
ridge_plots <- lapply(METRICS, function(m) plot_ridgeline(long_df, m))
patchwork::wrap_plots(ridge_plots, ncol = 2)
```

```{r timeseries, fig.height=10}
#| fig-cap: "Daily time series (lines break at missing days)."
ts_plots <- lapply(METRICS, function(m) plot_timeseries(long_df, m))
patchwork::wrap_plots(ts_plots, ncol = 2)
```

## Correlation and agreement — headline metrics

Pairwise comparisons on days where both devices recorded the metric. Deming regression slope/intercept test for proportional bias (slope ≠ 1 or intercept ≠ 0).

```{r agreement-table}
if (!is.null(agreement)) {
  agree_display <- agreement |>
    dplyr::mutate(
      metric = METRIC_LABELS[metric],
      dplyr::across(where(is.numeric), ~ round(.x, 3))
    )
  knitr::kable(agree_display, caption = "Pairwise agreement — headline metrics")
}
```

```{r corr-heatmap, fig.height=6}
#| fig-cap: "Pearson r heatmap (* p<0.05, ** p<0.01)."
if (!is.null(agreement)) plot_correlation_heatmap(agreement)
```

**Interpretation:** Whoop–Oura pairs dominate with r > 0.9. Withings shows weak RHR and sleep agreement with the other two, partly due to definitional differences (see next section).

### Resting HR — definition matters

The headline `rhr` metric uses each device's default: Whoop resting HR, Oura average resting HR, Withings **average sleeping HR**. We also compute `rhr_rest` — a like-for-like comparison using Whoop RHR, Oura **lowest** RHR, and Withings **minimum sleeping HR**.

```{r rhr-rest-agreement}
rhr_rest_agree <- run_all_agreement(wide_df, metrics = "rhr_rest")
if (!is.null(rhr_rest_agree)) {
  knitr::kable(
    rhr_rest_agree |>
      dplyr::mutate(metric = "Resting HR (like-for-like)", dplyr::across(where(is.numeric), ~ round(.x, 3))),
    caption = "RHR agreement — like-for-like definition (Whoop RHR / Oura lowest / Withings min sleeping HR)"
  )
}
```

```{r rhr-dumbbell, fig.height=8}
#| fig-cap: "Daily RHR spread — original definition (left panel concept) vs like-for-like."
p1 <- plot_paired_dumbbell(wide_df, "rhr")
p2 <- plot_paired_dumbbell(wide_df, "rhr_rest")
if (!is.null(p1) && !is.null(p2)) patchwork::wrap_plots(list(p1, p2), ncol = 1)
```

### Scatter, Bland–Altman, and difference-over-time

```{r scatter-ba-rhr, fig.height=10}
#| fig-cap: "RHR: scatter (Deming line) and Bland–Altman for each pair."
p1 <- plot_scatter_pair(wide_df, "rhr", "whoop", "oura", "Whoop", "Oura Ring 4")
p2 <- plot_bland_altman(wide_df, "rhr", "whoop", "oura", "Whoop", "Oura Ring 4")
p3 <- plot_scatter_pair(wide_df, "rhr", "whoop", "withings", "Whoop", "Withings ScanWatch")
p4 <- plot_bland_altman(wide_df, "rhr", "whoop", "withings", "Whoop", "Withings ScanWatch")
patchwork::wrap_plots(list(p1, p2, p3, p4), ncol = 2)
```

```{r diff-ts-rhr, fig.height=4}
#| fig-cap: "RHR daily difference (Oura − Whoop) over time — checks for drift."
plot_difference_timeseries(wide_df, "rhr", "whoop", "oura", "Whoop", "Oura Ring 4")
```

```{r rolling-rhr, fig.height=4}
#| fig-cap: "7-day rolling Pearson r for RHR (Whoop vs Oura)."
plot_rolling_correlation(wide_df, "rhr", "whoop", "oura", "Whoop", "Oura Ring 4")
```

```{r scatter-ba-sleep, fig.height=10}
#| fig-cap: "Sleep duration: scatter and Bland–Altman."
s1 <- plot_scatter_pair(wide_df, "sleep_min", "whoop", "oura", "Whoop", "Oura Ring 4")
s2 <- plot_bland_altman(wide_df, "sleep_min", "whoop", "oura", "Whoop", "Oura Ring 4")
s3 <- plot_scatter_pair(wide_df, "sleep_min", "whoop", "withings", "Whoop", "Withings ScanWatch")
s4 <- plot_bland_altman(wide_df, "sleep_min", "whoop", "withings", "Whoop", "Withings ScanWatch")
patchwork::wrap_plots(list(s1, s2, s3, s4), ncol = 2)
```

```{r scatter-ba-steps, fig.height=10}
#| fig-cap: "Steps: scatter and Bland–Altman (Whoop steps manually logged)."
t1 <- plot_scatter_pair(wide_df, "steps", "whoop", "oura", "Whoop", "Oura Ring 4")
t2 <- plot_bland_altman(wide_df, "steps", "whoop", "oura", "Whoop", "Oura Ring 4")
t3 <- plot_scatter_pair(wide_df, "steps", "oura", "withings", "Oura Ring 4", "Withings ScanWatch")
t4 <- plot_bland_altman(wide_df, "steps", "oura", "withings", "Oura Ring 4", "Withings ScanWatch")
patchwork::wrap_plots(list(t1, t2, t3, t4), ncol = 2)
```

```{r scatter-ba-hrv, fig.height=5}
#| fig-cap: "HRV: Whoop vs Oura only."
h1 <- plot_scatter_pair(wide_df, "hrv", "whoop", "oura", "Whoop", "Oura Ring 4")
h2 <- plot_bland_altman(wide_df, "hrv", "whoop", "oura", "Whoop", "Oura Ring 4")
patchwork::wrap_plots(list(h1, h2), ncol = 2)
```

## Sleep architecture (stage comparison)

All three devices report deep, REM, and light sleep. Stage detection algorithms differ, so moderate disagreement is expected even when total sleep agrees well.

```{r sleep-arch, fig.height=12}
#| fig-cap: "Stacked sleep-stage bars per night, faceted by device (most recent nights shown)."
plot_sleep_architecture(arch_df, n_nights = 14)
```

```{r stage-agreement}
stage_agree <- run_all_agreement(wide_df, metrics = SLEEP_STAGE_METRICS)
if (!is.null(stage_agree)) {
  knitr::kable(
    stage_agree |>
      dplyr::mutate(metric = METRIC_LABELS[metric], dplyr::across(where(is.numeric), ~ round(.x, 3))),
    caption = "Sleep-stage agreement (deep / REM / light minutes)"
  )
}
```

## Sleep timing agreement

Bedtime and wake time compared across devices. Values are expressed as decimal hours (bedtime: hours after noon; wake: hours after midnight) so they can be averaged and compared directly.

```{r timing-agreement}
timing_agree <- run_all_agreement(wide_df, metrics = c("bedtime_hr", "waketime_hr"))
if (!is.null(timing_agree)) {
  knitr::kable(
    timing_agree |>
      dplyr::mutate(metric = METRIC_LABELS[metric], dplyr::across(where(is.numeric), ~ round(.x, 3))),
    caption = "Sleep timing agreement"
  )
}
```

```{r timing-ts, fig.height=8}
#| fig-cap: "Bedtime (hours after noon) over the study window."
bt_plots <- lapply(c("whoop", "oura", "withings"), function(d) {
  sub <- long_df |> dplyr::filter(metric == "bedtime_hr",
    grepl(DEVICE_LABELS[[d]], device))
  if (nrow(sub) == 0) return(NULL)
  ggplot2::ggplot(sub, ggplot2::aes(x = date, y = value)) +
    ggplot2::geom_point(color = DEVICE_COLORS[[DEVICE_LABELS[[d]]]], size = 2) +
    ggplot2::geom_line(color = DEVICE_COLORS[[DEVICE_LABELS[[d]]]], linewidth = 0.7) +
    ggplot2::labs(title = DEVICE_LABELS[[d]], x = "Date", y = "Bedtime (h after noon)") +
    ggplot2::theme_minimal(base_size = 11)
})
patchwork::wrap_plots(Filter(Negate(is.null), bt_plots), ncol = 1)
```

## Respiratory rate and sleep efficiency

```{r resp-eff-agreement}
re_agree <- run_all_agreement(wide_df, metrics = c("resp_rate", "sleep_eff"))
if (!is.null(re_agree)) {
  knitr::kable(
    re_agree |>
      dplyr::mutate(metric = METRIC_LABELS[metric], dplyr::across(where(is.numeric), ~ round(.x, 3))),
    caption = "Respiratory rate and sleep efficiency agreement"
  )
}
```

Withings does not export respiratory rate in 2026; comparisons are Whoop–Oura for resp_rate and all three for sleep_eff.

## Sensitivity analysis

Re-runs headline-metric agreement **excluding QC-flagged days** (low steps, large day-jumps). If results are similar, conclusions are robust to outlier days.

```{r sensitivity}
if (!is.null(sensitivity)) {
  sens_display <- sensitivity |>
    dplyr::filter(metric %in% METRICS) |>
    dplyr::mutate(metric = METRIC_LABELS[metric], dplyr::across(where(is.numeric), ~ round(.x, 3))) |>
    dplyr::select(sample, metric, device_a, device_b, n, pearson_r, mean_bias, mae)
  knitr::kable(sens_display, caption = "Agreement: full sample vs QC-cleaned sample")
}
```

## Extended exploratory analyses

### Weekday vs weekend

```{r weekday-weekend}
knitr::kable(
  ww_stats |>
    dplyr::filter(metric %in% METRICS) |>
    dplyr::mutate(metric = METRIC_LABELS[as.character(metric)], mean = round(mean, 2)),
  caption = "Mean values on weekdays vs weekends"
)
```

### Rank consistency (three-way)

On days all three devices recorded a metric, do they rank nights similarly?

```{r rank-consistency}
rank_rhr   <- rank_consistency(wide_df, "rhr", c("whoop", "oura", "withings"))
rank_sleep <- rank_consistency(wide_df, "sleep_min", c("whoop", "oura", "withings"))
rank_steps <- rank_consistency(wide_df, "steps", c("whoop", "oura", "withings"))
rank_all <- dplyr::bind_rows(rank_rhr, rank_sleep, rank_steps)
if (!is.null(rank_all) && nrow(rank_all) > 0) {
  knitr::kable(
    rank_all |> dplyr::mutate(metric = METRIC_LABELS[metric], spearman = round(spearman, 3)),
    caption = "Rank correlation between devices (Spearman)"
  )
}
```

### Kendall's W (three-way concordance)

```{r kendall-w}
if (!is.null(kendall_tbl) && nrow(kendall_tbl) > 0) {
  knitr::kable(
    kendall_tbl |>
      dplyr::mutate(metric = METRIC_LABELS[metric], kendall_w = round(kendall_w, 3)),
    caption = "Kendall's W — three-way ranking agreement (1 = perfect, 0 = none)"
  )
}
```

### Cross-metric within device (RHR vs HRV)

```{r cross-metric}
cross_rhr_hrv <- cross_metric_correlation(long_df, "rhr", "hrv")
if (!is.null(cross_rhr_hrv)) {
  knitr::kable(
    cross_rhr_hrv |> dplyr::mutate(dplyr::across(where(is.numeric), ~ round(.x, 3))),
    caption = "Within-device Spearman correlation: RHR vs HRV"
  )
}
```

Negative correlation expected: lower RHR tends to co-occur with higher HRV (better recovery).

### Lagged effects (steps → next-day recovery)

```{r lagged}
lag_rhr  <- lagged_correlation(long_df, "steps", "rhr", lag_days = 1)
lag_hrv  <- lagged_correlation(long_df, "steps", "hrv", lag_days = 1)
lag_all  <- dplyr::bind_rows(lag_rhr, lag_hrv)
if (!is.null(lag_all) && nrow(lag_all) > 0) {
  knitr::kable(
    lag_all |> dplyr::mutate(dplyr::across(where(is.numeric), ~ round(.x, 3))),
    caption = "Lagged Spearman: today's steps vs tomorrow's RHR/HRV"
  )
}
```

### Withings intraday HR profile

```{r intraday-hr, fig.height=5}
#| fig-cap: "Average hourly HR from Withings intraday recordings (Pacific time)."
if (!is.null(intraday_hr)) plot_intraday_hr_profile(intraday_hr)
```

### Whoop workout log

```{r workouts}
if (!is.null(workouts) && nrow(workouts) > 0) {
  knitr::kable(
    workouts |>
      dplyr::select(date, activity_name, duration_min, activity_strain, average_hr_bpm) |>
      dplyr::mutate(dplyr::across(where(is.numeric), ~ round(.x, 1))),
    caption = "Whoop workouts during the study window"
  )
} else {
  cat("No Whoop workouts in the analysis window.")
}
```

## Limitations

- **Single participant**, ~33-day window — exploratory only, not population inference.
- **Whoop steps** manually typed; gaps and implausible lows reduce reliability.
- **Withings HRV** absent in 2026 export.
- **Withings sleep** stops after July 18.
- **RHR definitions differ** across devices unless using the like-for-like `rhr_rest` metric.
- Devices use **proprietary algorithms**; disagreement ≠ clinical inaccuracy without a reference standard.
- All timestamps in **`r TZ`**.

## Appendix

```{r session-info}
#| code-fold: true
sessionInfo()
```
